Reproducible Research 
=====================
***Praveen kumar***

## Impact of Severe Weather Events on Public Health and Economy in the United States

### Synonpsis  
In this report, we aim to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Basic settings

```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(ggplot2)
library(dplyr)
library(plyr)
require(gridExtra)
```
### Data Processing

```{r cache=TRUE}
stormData = read.csv("StormData.csv", header = TRUE)
dim(stormData)
head(stormData,5)
summary(stormData)
```

There are 902297 rows and 37 columns in total.

### Data Preprocessing

```{r}
# Trim the data set to required columns only
stormEvent <- stormData[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
    "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

# Create subset for Question 1 and Question 2

# Select data for Fatalities and injuries for Question 1
eventHealth <- subset(stormEvent, !stormEvent$FATALITIES == 0 & !stormEvent$INJURIES == 
    0, select = c(EVTYPE, FATALITIES, INJURIES))

# Select data for Property Damage and Crop Damage for Question 2
eventEconomic <- subset(stormEvent, !stormEvent$PROPDMG == 0 & !stormEvent$CROPDMG == 
    0, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

```


#### Impact on Public Health and Injuries

In this section, we check the number of **fatalities** and **injuries** that are caused by the severe weather events. We would like to get the first 10 most severe types of weather events.

```{r}
# Create separate data set for Injury and Fatalities Fatalities
eventHealth_Death <- aggregate(eventHealth$FATALITIES, by = list(eventHealth$EVTYPE), 
    FUN = sum)
# Give proper name for columns
colnames(eventHealth_Death) <- c("EVENTTYPE", "FATALITIES")

# Injury
eventHealth_Inj <- aggregate(eventHealth$INJURIES, by = list(eventHealth$EVTYPE), 
    FUN = sum)
# Give column name
colnames(eventHealth_Inj) <- c("EVENTTYPE", "INJURIES")

# Let's reorder 2 dataset and filter top 10 events for both dataset
eventHealth_Death <- eventHealth_Death[order(eventHealth_Death$FATALITIES, decreasing = TRUE), 
    ][1:10, ]

eventHealth_Inj <- eventHealth_Inj[order(eventHealth_Inj$INJURIES, decreasing = TRUE), 
    ][1:10, ]
```

#### Impact on Economy

We will convert the **property damage** and **crop damage** data into comparable numerical forms according to the meaning of units described in the code book ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}
# select required entries for economy
eventEconomic <- subset(eventEconomic, eventEconomic$PROPDMGEXP == "K" | eventEconomic$PROPDMGEXP == 
    "k" | eventEconomic$PROPDMGEXP == "M" | eventEconomic$PROPDMGEXP == "m" | 
    eventEconomic$PROPDMGEXP == "B" | eventEconomic$PROPDMGEXP == "b")

eventEconomic <- subset(eventEconomic, eventEconomic$CROPDMGEXP == "K" | eventEconomic$CROPDMGEXP == 
    "k" | eventEconomic$CROPDMGEXP == "M" | eventEconomic$CROPDMGEXP == "m" | 
    eventEconomic$CROPDMGEXP == "B" | eventEconomic$CROPDMGEXP == "b")

# Convert ecnomic values to number
eventEconomic$PROPDMGEXP <- gsub("m", 1e+06, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- gsub("k", 1000, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- gsub("b", 1e+09, eventEconomic$PROPDMGEXP, ignore.case = TRUE)
eventEconomic$PROPDMGEXP <- as.numeric(eventEconomic$PROPDMGEXP)
eventEconomic$CROPDMGEXP <- gsub("m", 1e+06, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- gsub("k", 1000, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- gsub("b", 1e+09, eventEconomic$CROPDMGEXP, ignore.case = TRUE)
eventEconomic$CROPDMGEXP <- as.numeric(eventEconomic$CROPDMGEXP)
eventEconomic$PROPDMGEXP <- as.numeric(eventEconomic$PROPDMGEXP)

# then sum the damages by each event type
eventEconomic$TOTALDMG <- (eventEconomic$CROPDMG * eventEconomic$CROPDMGEXP) + 
    (eventEconomic$PROPDMG * eventEconomic$PROPDMGEXP)

eventEconomic <- aggregate(eventEconomic$TOTALDMG, by = list(eventEconomic$EVTYPE), 
    FUN = sum)

colnames(eventEconomic) <- c("EVTYPE", "TOTALDMG")

# Rank the event type by highest damage cost and take top 10 columns
eventEconomic <- eventEconomic[order(eventEconomic$TOTALDMG, decreasing = TRUE), 
    ]
eventEconomic <- eventEconomic[1:10, ]

```

### Results

The following is a pair of graphs of total fatalities and total injuries affected by these severe weather events. 

```{r}

# plot top 10 events for fatalities and injuries

# Plot Fatalities and store at Death_plot
Death_plot <- ggplot() + geom_bar(data = eventHealth_Death, aes(x = EVENTTYPE, 
    y = FATALITIES, fill = interaction(FATALITIES, EVENTTYPE)), stat = "identity", 
    show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    xlab("Harmful Events") + ylab("No. of fatailities") + ggtitle("Top 5 weather events causing fatalities") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Plot injuries and store at variable Inj_plot
Inj_plot <- ggplot() + geom_bar(data = eventHealth_Inj, aes(x = EVENTTYPE, y = INJURIES, 
    fill = interaction(INJURIES, EVENTTYPE)), stat = "identity", show.legend = F) + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + xlab("Harmful Events") + 
    ylab("No. of Injuries") + ggtitle("Top 5 weather events causing Injuries") + 
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Draw two plots generated above dividing space in two columns

grid.arrange(Death_plot, Inj_plot, ncol = 2)

```

Based on the above ggplots, we find that **excessive heat** and **tornado** cause most fatalities; **tornato** causes most injuries in the United States from 1995 to 2011.

The following is a graph of total property damage affected by these severe weather events. 

```{r}
# Now plot the graph
ggplot() + geom_bar(data = eventEconomic, aes(x = EVTYPE, y = TOTALDMG, fill = interaction(TOTALDMG, 
    EVTYPE)), stat = "identity", show.legend = F) + theme(axis.text.x = element_text(angle = 30, 
    hjust = 1)) + xlab("Event Type") + ylab("Total Damage")
```

### Conclusion  

From these data, we found that **excessive heat** and **tornado** are most harmful with respect to population health, while **flood**, **drought**, and **hurricane/typhoon** have the greatest economic consequences.
